When does vapor pressure deficit drive or reduce evapotranspiration?\footnote{Primary Research Article}

### 2018-04-03 doc/paper/gcb_sub_vpd_et_paper.tex
successfully generated diff for Pierre

% 11/23/2015
\documentclass[draft,linenumbers]{agujournal}
% \draftfalse
\usepackage{makecell} \usepackage{multirow} %AM for tables
\usepackage{booktabs} \drafttrue

\journalname{Agricultural and Forest Meteorology: When does VPD drive
  or reduce ET?}

@@

\title{When does vapor pressure deficit drive or reduce
  evapotranspiration?}

\authors{Adam Massmann\affil{a}, Pierre Gentine\affil{a}, Changjie
  Lin\affil{a,b}}

\affiliation{a}{Department of Earth and Environmental Engineering,
  Columbia University, New York, NY 10027} \affiliation{b}{State Key
  Laboratory of Hydroscience and Engineering, Department of Hydraulic
  Engineering, Tsinghua University, Beijing, CN 100084}

@@

\begin{abstract}
  Increasing vapor pressure deficit increases atmospheric demand for
  water, and vapor pressure deficit is expected to rise with
  increasing CO$_2$. While increased evapotranspiration (ET) in
  response to increased atmospheric demand seems intuitive, plants are
  capable of reducing ET in response to increased VPD by closing their
  stomata, in an effort to conserve water. Here we examine which
  effect dominates response to increasing VPD: atmospheric demand and
  increases in ET, or plant physiological response (stomata closure)
  and decreases in ET. We use Penman-Monteith, combined with
  semi-empirical optimal stomatal regulation theory and underlying
  water use efficiency, to develop a theoretical framework for
  understanding how ET responds to increases in VPD.

  The theory suggests that for most environmental conditions and plant
  types, plant physiological response dominates and ET decreases with
  increasing VPD. Plants that are evolved or bred to prioritize
  primary production over water conservation (e.g. crops) exhibit a
  higher likelihood of atmospheric demand-driven response (ET
  increasing). However for forest, grass, savannah, and shrub plant
  types, ET more frequently decreases than increases with rising
  VPD. This work serves as an example of the utility of
  our simplified framework for disentangling land-atmosphere feedbacks
  at multiple scales, including the characterization of ET response in
  an atmospherically drier, enriched CO$_2$ world.

@@

The opposing effects of increased atmospheric demand and higher
stomatal stress lead to two possible perspectives for how
evapotranspiration (ET) responds to shifts in VPD. The first, a
hydrometeorological perspective, is that higher VPD increases
atmospheric demand for water from the land surface, and this drives an
increase in evapotranspiration (ET). This perspective is particularly
relevant because potential evapotranspiration (PET), which is used in
many drought indices and hydrometeorological studies
\citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies
changes in atmospheric demand and fails to account for ecosystem
response \citep{Swann_2016}. In reality, plants' stomata have evolved
to optimally regulate the exchange of water and carbon, and tend to
partially close in response to increased atmospheric dryness
\citep{Farquhar_1978, Ball_1987, Leuning_1990, MEDLYN_2011}. This
leads to a plant physiology perspective, in which an increase in VPD,
particularly in well-watered soil conditions, may actually correspond
to a decrease in ET because of stomatal closure
\citep[e.g.][]{Rigden_2017}.  In other words, the question ``When does
VPD drive or reduce ET?'' can be related to whether plant regulation
or atmospheric demand dominates ET response.

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with
atmospheric drying then soil moisture will be better conserved. This
would seem a sensible evolutionary strategy to cope with aridity. If
stomata were fully passive \citep [similar to soil pores,
e.g. ][]{Or_2013}, increased atmospheric aridity would strongly reduce
soil moisture \citep{Berg_2017}. In turn, this would further increase
aridity as low soil moisture levels increase the Bowen ratio, and
cause increased temperature and atmospheric drying
\citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
  Salvucci_2013, Gentine_2016, Berg_2016}. This however would not seem
to be a sensible strategy for plants from an evolutionary
standpoint. As a counterpoint, one may argue that increases in ET with
increasing VPD could increase the likelihood of precipitation
\citep[e.g.,][]{Findell_2011}. However, increases in ET do not always
guarantee an increased likelihood of precipitation, which depending on
environmental conditions could cause a decrease in the likelihood
of precipitation \citep[][]{Gentine_2013}. Furthermore, any increases
in precipitation are likely to be non-local, such that plants giving
up water to the atmosphere are not guaranteed to reap the benefits of
water returned from the atmosphere to the soil. Using this logic, from
an ecosystem evolutionary perspective, water stored in soil is worth
much more than the chance of water returned as precipitation.

We can use intuition about plant water conservation strategy to
hypothesize about ET response to changes in VPD. Plants and ecosystems
that evolved to conserve water, such as arid shrubs or savannah,
should be more likely to reduce ET with increasing VPD, and plants
that have evolved or have been engineered to care little about water,
such as crops, will be more likely to increase ET with increasing
VPD. Atmospheric conditions must matter as well. At the ecosystem
scale, there are limits to plant water conservation strategies. As
atmospheric demand for water (VPD) increases, ecosystems should begin
to reach their water conservation limits and might not be able to
entirely limit ET flux to the atmosphere. At this stage any further
increase in VPD will most likely drive a (limited) increase in ET,
because the increase in atmospheric demand for water overwhelms the
limited plant response to conserve water.

The objective of the present manuscript is to use reasonable
approximations as a tool to develop intuition for plant response to
atmospheric drying and evaluate the VPD dependence of ET. This
intuition will aid interpretation of observations, full complexity
models, and facilitate the disentanglement of complex land-atmosphere
feedbacks. In the past, similar approaches were used to understand
interactions between stomatal conductance, evapotranspiration and the
environment \citep[e.g.,][]{Jarvis_1986, Mcnaughton_1991}. However, at
the time researchers' understanding of the form of VPD's effect on
plant physiology was limited, so they could not explore the
sensitivity of ET to VPD, including VPD's effect on stomatal
conductance and plant function.

Recent results have drastically improved our understanding of VPD's
impact on physiology. \citet{MEDLYN_2011} developed a model for
leaf-scale stomatal conductance ($g_s$), including VPD response, by
combining an optimal photosynthesis theory \citep{Farquhar_1980,
  Katul_2010} with an empirical approach, and extended use of this
model to the ecosystem scale in \citet{Medlyn_2017}. Additionally,
\citet{Zhou_2014} demonstrated that a quantity underlying water use
efficiency $\left(uWUE = \frac{GPP\; \sqrt{VPD}}{ET}\right)$ properly
captures a constant relationship between GPP, ET, and VPD over a
diurnal cycle. uWUE is also remarkably well conserved in the growing
season across space and time, within a PFT \citep{Zhou_2015}. These
results provide novel tools to examine how ET varies in response to
VPD, while explicitly including VPD's effect on stomatal conductance
for the first time. In this manuscript, we leverage these tools to
derive the theoretical response of ET to VPD using a Penman Monteith
framework. Our theory is validated and tested at multiple
eddy-covariance stations spanning various climates and plant
functional types.

\section{Materials and Methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the
FLUXNET2015 database (data available at \sloppy
https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/ \sloppy),
including all sites with at least four years of data and observations
of the variables described in the methods section (Section
\ref{methods}). Seventy-three sites met these requirements, and were
grouped into nine plant functional types (PFT) according to the
International Geosphere-Biosphere Programme vegetation classification
scheme \citep{Loveland_1999}: cropland (CRO), grass (GRA), deciduous
broadleaf forest (DBF), evergreen broadleaf forest (EBF), evergreen
needleleaf forest (ENF), mixed forest (MF), closed shrub (CSH),
savannah (SAV), and woody savannah (WSA) (site locations in Figure
\ref{map_fig}).

@@

The purpose of this study is to examine ecosystem response to
atmospheric drying, focusing on the growing season. To accomplish
this, we filter and quality control the data using a similar procedure
as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data,
  according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the
  average Gross Primary Productivity (GPP) exceeds 10\% of the
  observed 95th percentile of GPP for a given site. GPP is calculated
  using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues
  with rain interception and sensor saturation at high relative
  humidity (\cite{Medlyn_2017}).
\end{itemize}
Additionally, as in \citet{Lin_2018}, we restrict data to the daytime,
which is identified when downwelling shortwave radiation is greater
than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W
m$^{-2}$. To reduce the chance of sensor saturation at high relative
humidity, we remove all time steps for which VPD is less than .01 kPa,
and to reduce errors at low windspeeds we remove all periods with wind
magnitudes less than 0.5 m s$^{-1}$. Timesteps with negative observed
GPP or ET are also removed, and we aggregate half hourly data to
hourly averages to reduce noise \citep{Lin_2018}. After these quality
control procedures, 434,765 upscaled hourly observations remain.

@@

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \small
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\ 
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{s-leaf}$  & leaf-scale stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citet{Medlyn_2017} \\
    $g_{1-leaf}$  & leaf-scale slope parameter & Pa$^{0.5}$
                                   & \citet{MEDLYN_2011} \\
    $g_{1}$  & ecosystem-scale slope parameter & Pa$^{0.5}$ & \citet{MEDLYN_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\
    \hline
  \end{tabular}
\end{table}

\citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$)
by combining an optimal photosynthesis theory \citep{Cowan_1977} with an empirical approach, which describes the
dependence of $g_s$ to VPD. This resulted in the following model for
leaf-scale stomatal conductance:

\begin{linenomath*}
  \begin{equation}
    g_{s-leaf} = g_0 + 1.6 \left(1 +
      \frac{g_{1-leaf}}{\sqrt{VPD}}\right) \frac{A}{c_a},
    \label{leaf_medlyn}
  \end{equation}
\end{linenomath*}
where $g_{1-leaf}$ is a leaf-scale ``slope'' parameter, A is the net
CO$_2$ assimilation rate, and $c_a$ is the atmospheric CO$_2$
concentration at the leaf surface. \cite{MEDLYN_2011} relate the slope
parameter ($g_{1-leaf}$) to physical parameters as:
\begin{linenomath*}
  \label{slope}
  \begin{equation}
    g_{1-leaf} = \sqrt{\frac{3 \, \Gamma^* \, \lambda}{1.6}},\footnote{Note this expression has units of of (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to Pa$^{1/2}$ using the ideal gas law.}
  \end{equation}
\end{linenomath*}

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis
(without dark respiration), and $\lambda$ is the marginal water cost
of leaf carbon
($\frac{\partial \; \text{transpiration}}{\partial \; A}$). So,
$g_{1-leaf}$ is a leaf-scale term reflecting the trade-off of water for
carbon uptake. The higher $g_{1-leaf}$, the more open the stomata and
the more they release water in exchange for carbon.

The Medlyn model for stomatal conductance has been shown to behave
very well across PFTs \citep[][]{Lin_2015}, and has been successfully
adopted to ecosystem scale analysis in \citet{Medlyn_2017}. In units
of m s$^{-1}$, the ecosystem scale stomatal conductance is:

\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \,T}{P} \; 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{GPP}{c_a},
    \label{medlyn}
  \end{equation}
\end{linenomath*}

where GPP is the ecosystem scale gross primary production, and $g_1$
is an ecosystem scale analogue to $g_{1-leaf}$. We solve for $g_1$
following \citet{Medlyn_2017} (Equation 5), and take the median $g_1$
value to be representative of each PFT (Table \ref{pft}).

\begin{table}
  \caption{Plant functional types, their abbreviation, calculated
    Medlyn coefficient, and calculated uWUE. uWUE from
    \citet{Zhou_2015}, including the observed standard deviation is
    shown for comparison. Note that uWUE from \citet{Zhou_2015} is
    calculated from a different set of sites, and that units are
    converted such that the quantities work with Equations 1-8 and the
    variables defined Table \ref{definitions}.}
  \small
  \label{pft}
  \centering
  \begin{tabular}{l c c @{\qquad} c c}
    \hline
    \multirow{2}[3]{*}{Abbreviation} & \multirow{2}[3]{*}{PFT} & \multirow{2}[3]{*}{$g_1$ (Pa$^{0.5}$)} & \multicolumn{2}{c}{uWUE ($\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET])}  \\
    \cmidrule{4-5}

    \hline
    \input{pft_params}
    \hline
  \end{tabular}
\end{table}

While Equation 3 can be used in PM (Equation 1), it will make
analytical work with the function intractable because $GPP$ is
functionally related to ET itself. Additionally, a perturbation to VPD
should induce a physiological plant response that will alter GPP and
cause an indirect change in stomatal conductance, in addition to the
direct effect of VPD in Equation \ref{medlyn}. Therefore, in order to
derive the response of ET to VPD, we must account for the functional
relationship between GPP, ET, and VPD, and its effect on stomatal
conductance. We can use aforementioned semi-empirical results of
\citet{Zhou_2015} as a tool to approach this
problem. \citet{Zhou_2015}, showed that  underlying Water Use
Efficiency (uWUE):

\begin{linenomath*}
  \begin{equation}
    uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET}
    \label{uwue}
  \end{equation}
\end{linenomath*}

is relatively constant across time and moisture conditions within a
plant functional type, and correctly captures a constant relationship
between GPP, ET and VPD over a diurnal cycle \citep{Zhou_2014}. We can
use uWUE to remove the $GPP$ dependence from $g_s$ in a way that makes
the PM equation analytically tractable:

@@

Plugging Equation \ref{new_g_s} into Equation \ref{orig_pen} and
rearranging gives a new explicit expression for PM, in which
dependence on $GPP$ is removed:

@@

With Equation \ref{et} we have derived for the first time, using
recent results \citep[][]{MEDLYN_2011, Zhou_2014, Zhou_2015,
  Medlyn_2017}, ET explicitly as function of environmental variables
and two plant-specific constants, the slope parameter ($g_1$), and
uWUE, both reflecting water conservation strategy. The slope parameter
is related to the willingness of stomata to trade water for CO$_2$ and
to keep stomata open. uWUE is a semi-empirical ecosystem-scale
constant related to how WUE changes with VPD (specifically
$VPD^{-1/2}$). It is also roughly proportional to physical constants:

@@

Given eddy-covariance FLUXNET2015 data (Section \ref{data}), every
term in our new version of PM (Equation \ref{et}) is observed, except
for uWUE, which we fit by calculating its expectation, given the model
and FLUXNET2015 data.

However, eddy-covariance data are inherently noisy so we include a
measure of uncertainty in our analysis. To account for observational
error, as well as model uncertainty (e.g. temporal and spatial
variations of uWUE and $g_1$), we introduce an uncertainty parameter
$\sigma$ modifying uWUE:

@@

Thus, with this uncertainty analysis we can evaluate departure from
our theory in observations, as a departure of $\sigma$ from unity. The
variability of $\sigma$ across sites and time provides a measure of
uncertainty in our model, assumptions, as well as in the FLUXNET2015
observations themselves. The variability of $\sigma$ then propagates
through any uncertainty to our partial derivative of Equation
\ref{et_sigma} with respect to VPD:

@@

With Equation \ref{d_et} we have an analytical framework for ecosystem
response to atmospheric demand perturbations with environmental
conditions held fixed. There are a few subtleties to taking the
derivative in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$)
and $VPD$ are functionally related, so while taking the derivative we
evaluate
$\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
{\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
\Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
\frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
    fixed}}$. $RH$ and $e_s$ are assumed to be approximately
independent, which is supported by the data (not shown).

>>>>

% 11/23/2015
\documentclass[draft,linenumbers]{agujournal}
\draftfalse \usepackage{makecell} \usepackage{multirow} %AM for tables
\usepackage{booktabs} \drafttrue

\journalname{Global Change Biology: When does VPD drive or reduce ET?}

@@

\title{When does vapor pressure deficit drive or reduce
evapotranspiration?\footnote{Primary Research Article}}

\authors{Adam Massmann\affil{1}, Pierre Gentine\affil{1}, Changjie
Lin\affil{1,2}}

\affiliation{1}{Department of Earth and Environmental Engineering,
Columbia University, New York, NY 10027} \affiliation{2}{State Key
Laboratory of Hydroscience and Engineering, Department of Hydraulic
Engineering, Tsinghua University, Beijing, CN 100084}

@@

\begin{abstract} Increases to vapor pressure deficit (VPD) and
atmospheric demand for water are expected with rising atmospheric
CO$_2$. While increased evapotranspiration (ET) in response to
increased atmospheric demand seems intuitive, plants are capable of
reducing ET in response to increased VPD by closing their stomata, in
an effort to conserve water. Here we examine which effect dominates
response to increasing VPD: atmospheric demand and increases in ET, or
plant physiological response (stomata closure) and decreases in ET. We
use Penman-Monteith, combined with semi-empirical optimal stomatal
regulation theory and underlying water use efficiency, to develop a
theoretical framework for understanding how ET responds to increases
in VPD.  The theory suggests that for most environmental conditions
and plant types, plant physiological response dominates and ET
decreases with increasing VPD. Plants that are evolved or bred to
prioritize primary production over water conservation (e.g. crops)
exhibit a higher likelihood of atmospheric demand-driven response (ET
increasing). However for forest, grass, and shrub plant types, ET more
frequently decreases than increases with rising VPD. Therefore, we may
expect reductions in ET due to CO$_2$-induced increases in VPD. This
work serves as an example of the utility of our simplified framework
for disentangling land-atmosphere feedbacks at multiple scales,
including the characterization of ET response in an atmospherically
drier, enriched CO$_2$ world.

@@

The opposing effects of increased atmospheric demand and higher
stomatal stress lead to two possible perspectives for how
evapotranspiration (ET) responds to shifts in VPD. The first, a
hydrometeorological perspective, is that higher VPD increases
atmospheric demand for water from the land surface, and this drives an
increase in evapotranspiration (ET). This perspective is particularly
relevant because potential evapotranspiration (PET), which is used in
many drought indices and hydrometeorological studies
\citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies
changes in atmospheric demand and fails to account for ecosystem
response \citep{Swann_2016}. In reality, plants' stomata have evolved
to optimally regulate the exchange of water and carbon, and tend to
partially close in response to increased atmospheric dryness
\citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant
physiology perspective, in which an increase in VPD, particularly in
well-watered soil conditions, may actually correspond to a decrease in
ET because of stomatal closure.  In other words, the question ``When
does VPD drive or reduce ET?'' can be related to whether plant
regulation or atmospheric demand dominates ET response.

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If plant response reduces ET with atmospheric
drying then soil moisture will be better conserved. This would seem a
sensible evolutionary strategy to cope with aridity. If stomata were
fully passive \citep [similar to soil pores, e.g. ][]{Or_2013},
increased atmospheric aridity would strongly reduce soil moisture
\citep{Berg_2017}. In turn, this would further increase aridity as low
soil moisture levels increase the Bowen ratio, and cause increased
temperature and atmospheric drying \citep[][]{Bouchet_1963,
Morton_1965, Brutsaert_1999, Ozdogan_2006, Salvucci_2013,
Gentine_2016, Berg_2016}. This however would not seem to be a sensible
strategy for plants from an evolutionary standpoint.

We can use intuition about plant water conservation strategy to
hypothesize about ET response to changes in VPD. Plants that evolved
to conserve water, such as arid shrubs, should be more likely to
reduce ET with increasing VPD, and plants that have evolved or have
been engineered to care little about water, such as crops, will be
more likely to increase ET with increasing VPD. Atmospheric conditions
must matter as well. At the ecosystem scale, there are limits to plant
water conservation strategies. As atmospheric demand for water (VPD)
increases, ecosystems should begin to reach their water conservation
limits and might not be able to entirely limit ET flux to the
atmosphere. At this stage any further increase in VPD will most likely
drive a (limited) increase in ET, because the increase in atmospheric
demand for water overwhelms the limited plant response to conserve
water.

The objective of the present manuscript is to evaluate the VPD
dependence of ET, in non-extreme soil drought conditions. The goal of
this paper is to use reasonable approximations as a tool to develop
intuition for plant response to atmospheric drying. This intuition
will aid interpretation of observations and full complexity models. In
order to quantify plant response to perturbations in VPD, we apply a
Penman-Monteith framework to derive the theoretical response of ET to
VPD. The model is then validated and tested at multiple
eddy-covariance stations spanning various climates and plant
functional types.

\section{Materials and Methods}
\label{methods}
\subsection{Data}
\label{data} We use both meteorological and eddy-covariance data from
the FLUXNET2015 database, including all sites with at least four years
of data (data available at \sloppy
https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/ \sloppy). Each
site's plant functional type (PFT) was classified using the
International Geosphere-Biosphere Programme vegetation classification
scheme \citep{Loveland_1999}. The physical constants used in the
stomatal conductance model in the Methods section (Section
\ref{methods}) are only published for six plant functional types
(PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreen
broadleaf forests (EBF), evergreen needleleaf forest (ENF), grass
(GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 62
sites with these plant functional types, and their location is shown
in Figure \ref{map_fig}.

\begin{table}
  \caption{Plant functional types, their abbreviation, Medlyn
coefficient (from \citet{Lin_2015}, via \citet{Franks_2017}), and
calculated uWUE. uWUE from \citet{Zhou_2015}, including the observed
standard deviation is shown for comparison. Note that uWUE from
\citet{Zhou_2015} is calculated from a different set of sites, and
that units are converted such that the quantities work with Equations
1-8 with the variables defined Table \ref{definitions}.}  \small
\label{pft} \centering
\begin{tabular}{l c c @{\qquad} c c} \hline
\multirow{2}[3]{*}{Abbreviation} & \multirow{2}[3]{*}{PFT} &
\multirow{2}[3]{*}{$g_1$ (Pa$^{0.5}$)} & \multicolumn{2}{c}{uWUE
($\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET])} \\ \cmidrule{4-5}

  \hline DBF & deciduous broadleaf forest & 140.7 & 2.63 & 3.12 $\pm$
0.52 \\ EBF & evergreen broadleaf forest & 130.3 & 2.56 & N/A \\ ENF &
evergreen needleleaf forest & 74.3 & 3.45 & 3.30 $\pm$ 0.91 \\ CRO &
cropland & 183.1 & 2.27 & 3.80 $\pm$ 1.01 \\ CSH & closed shrub &
148.6 & 1.66 & 2.18 $\pm$ 0.44 \\ GRA & grassland (assumed C3) & 166.0
& 1.92 & 2.68 $\pm$ 0.61 \\ \hline
\end{tabular}
\end{table}

@@

The purpose of this study is to examine ecosystem response to
atmospheric drying, focusing on the growing season. To accomplish
this, we filter and quality control the data using a similar procedure
as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data,
according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the
average Gross Primary Productivity (GPP) exceeds 10\% of the observed
95th percentile of GPP for a given site. GPP is calculated using the
nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues
with rain interception and sensor saturation at high relative humidity
(\cite{MEDLYN_2011}).
\end{itemize} Additionally, as in \citet{Lin_2018}, we restrict data
to the daytime, which is identified when downwelling shortwave
radiation is greater than 50 W m$^{-2}$ and sensible heat flux is
greater than 5 W m$^{-2}$. To reduce the chance of sensor saturation
at high relative humidity, we remove all time steps for which VPD is
less than .01 kPa, and to reduce errors at low windspeeds we remove
all periods with wind magnitudes less than 0.5 m s$^{-1}$. Timesteps
with negative observed GPP or ET are also removed, and we aggregate
half hourly data to hourly averages to reduce noise
\citep{Lin_2018}. After these quality control procedures, 335,939
upscaled hourly observations remain.

@@

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
values are calculated, if applicable.}
  \label{definitions} \centering \small
\begin{tabular}{l c c c} \hline Variable & Description & Units &
Citation \\ \hline $e_s$ & saturation vapor pressure & Pa & - \\ $T$ &
temperature & K & - \\ $P$ & pressure & Pa & - \\ $\Delta$ &
$\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\ $R_{net}$ & net
radiation at land surface minus ground heat flux & W m$^{-2}$ & - \\
$g_a$ & aerodynamic conductance & m s$^{-1}$ &
\citet{Shuttleworth_2012} \\ $\rho_a$ & air density & kg m$^{-3}$ & -
\\ $c_p$ & specific heat capacity of air at constant pressure & J
K$^{-1}$ kg$^{-1}$ & - \\ $VPD$ & vapor pressure deficit & Pa & - \\
$\gamma$ & psychometric constant & Pa K$^{-1}$ & - \\ $g_s$ & stomatal
conductance & m s$^{-1}$ & \citet{MEDLYN_2011} \\ $g_{l-s}$ &
leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$ &
\citet{MEDLYN_2011} \\ $R$ & universal gas constant & J mol$^{-1}$
K$^{-1}$ & - \\ $R_{air}$ & gas constant of air & J K$^{-1}$ kg$^{-1}$
& - \\ $LAI$ & leaf area index & -& - \\ $\sigma$ & uncertainty
parameter & -& - \\ $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$
mol$^{-1}$ air& - \\ $\lambda$ & marginal water cost of leaf carbon &
mol H$_2$O mol$^{-1}$ CO$_2$ & - \\ $\Gamma$ & CO$_2$ compensation
point & - & - \\ $\Gamma^*$ & CO$_2$ compensation point without dark
respiration & - & - \\ \hline
\end{tabular}
\end{table}

\citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$)
by combining an optimal photosynthesis theory \citep{Farquhar_1980,
Katul_2010} with an empirical approach, which describes the dependence
of $g_s$ to VPD. This resulted in the following model for leaf-scale
stomatal conductance:

\begin{linenomath*}
  \begin{equation} g_{l-s} = g_0 + 1.6 \left(1 +
\frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_a}
  \end{equation}
\end{linenomath*} Where $g_1$ is a leaf-scale ``slope'' parameter, A
is the net CO$_2$ assimilation rate, and $c_a$ is the atmospheric
CO$_2$ concentration at the leaf surface. \cite{MEDLYN_2011} relate
the slope parameter ($g_1$) to physical parameters as:
\begin{linenomath*}
  \label{slope}
  \begin{equation} g_1 = \sqrt{\frac{3 \, \Gamma^* \,
\lambda}{1.6}},\footnote{Note this expression has units of of (mmol
mol$^{-1}$)$^{1/2}$, but this can be converted to Pa$^{1/2}$ using the
ideal gas law.}
  \end{equation}
\end{linenomath*}

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis
(without dark respiration), and $\lambda$ is the marginal water cost
of leaf carbon ($\frac{\partial \; \text{transpiration}}{\partial \;
A}$). So, $g_1$ is a leaf-scale term reflecting the tradeoff of water
for carbon uptake and growth. The higher $g_1$, the more open the
stomata and the more they release water in exchange for carbon.

The Medlyn model for stomatal conductance has been shown to behave
very well across PFTs \citep[][]{Lin_2015}, and can be adapted to the
ecosystem scale by multiplying $g_{l-s}$ by leaf area index (LAI) and
converting units to m s$^{-1}$ with the ideal gas law:

\begin{linenomath*}
  \begin{equation} g_s = \text{LAI} \frac{R \,T}{P} \left( g_0 + 1.6
\left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_a}\right)
      \label{medlyn}
  \end{equation}
\end{linenomath*}

While Equation 3 can be used in PM (Equation 1), it will make
analytical work with the function intractable because $A$, net CO$_2$
assimilation rate, is functionally related to ET itself. To remove the
dependence of ET on $A$ we use fundamental semi-empirical results of
\citet{Zhou_2015}, who showed that the underlying Water Use Efficiency
(uWUE):

\begin{linenomath*}
  \begin{equation} uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET}
    \label{uwue}
  \end{equation}
\end{linenomath*} is relatively constant across time and moisture
conditions within a plant functional type. If, following
\citet{Lin_2015}, we approximate $g_0$ as $0$ (i.e. we neglect
cuticular and epidermal losses - a reasonable assumption except in
very dry conditions), we can use uWUE to remove the $A$ dependence
from $g_s$ in a way that makes the PM equation analytically tractable:

@@

Note that uWUE is fit on the ecosystem scale as in \citet{Zhou_2015},
so GPP in Equation \ref{uwue} is really $A \cdot \text{LAI}$. This
leads to the cancellation of LAI in the step from Equation
\ref{medlyn} to Equation \ref{new_g_s}. Plugging Equation
\ref{new_g_s} into Equation \ref{orig_pen} and rearranging gives a new
explicit expression for PM, in which dependence on $A$ is removed:

@@

With Equation \ref{et}, ET is explicitly a function of environmental
variables and two plant-specific constants, the slope parameter
($g_1$), and uWUE, both reflecting water conservation strategy. The
slope parameter is a leaf-scale term related to the willingness of
stomata to trade water for CO$_2$ and to keep stomata open. uWUE is a
semi-empirical ecosystem-scale constant related to how WUE changes
with VPD (specifically $VPD^{-1/2}$). Is is also roughly proportional
to physical constants:

@@

Given eddy-covariance FLUXNET2015 data (Section \ref{data}), every
term in our new version of PM (Equation \ref{et}) is observed, except
for the two parameters $g_1$, and uWUE. $g_1$ has been measured and
reported for the PFTs considered here \citep{Lin_2015, Franks_2017},
and we fit uWUE by calculating its expectation, given the model and
FLUXNET2015 data.

However, eddy-covariance data are inherently noisy so we include a
measure of uncertainty in our analysis. To account for observational
error, as well as model uncertainty (e.g. temporal and spatial
variations of uWUE), we introduce an uncertainty parameter $\sigma$
modifying uWUE:

@@

Thus, with this uncertainty analysis we can evaluate departure from
our theory in observations, as a departure of $\sigma$ from unity. The
variability of $\sigma$ across sites and time provides a measure of
uncertainty in our model, assumptions, as well as in the FLUXNET2015
observations themselves. The variability of $\sigma$ then propagates
through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

With Equation \ref{d_et} we have an analytical framework for ecosystem
response to atmospheric demand perturbations, with an implicit
assumption of non-extreme drought conditions. There are a few
subtleties to taking the derivative in Equation \ref{d_et}: $\Delta$
($\frac{d e_{s}}{d T}$) and $VPD$ are functionally related, so while
taking the derivative we evaluate $\frac{\partial \; ET}{\partial \;
VPD} = \frac{\partial \; ET} {\partial \; e_s} \frac{\partial \;
e_s}{\partial \; VPD} \Big|_{\text{RH fixed}} + \frac{\partial \;
ET}{\partial \; RH} \frac{\partial \; RH}{\partial \; VPD}
\Big|_{\text{$e_s$ fixed}}$. $RH$ and $e_s$ are assumed to be
approximately independent, which is supported by the data (not shown).

We note one final comment on our derivation which is relevant for
drought indices. If we approximate $c_a$ at a global mean CO$_2$
concentration, then the RHS of Equation \ref{et} is fully defined
using commonly available weather station data and the constants
published in \citet{Zhou_2015} and \citet{Lin_2015}. This makes
Equation \ref{et} a useful alternative to PET in drought indices and
hydrometeorological analysis for vegetated surfaces. Equation \ref{et}
better reflects the physics of water exchange at the land surface and
would only require fitting of uWUE and $g_1$, and can also account for
changes in CO$_2$ concentration, contrary to typical drought indices
such as the Palmer Drought Severity Index (PDSI) \citep{Swann_2016}.
